---
title: "Missed reproductive events can reduce accuracy of uncertainty
estimates in matrix population models"
author: "Emily Simmonds"
date: "2023-03-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# load packages
library(tidyverse)

# source script

source("./Scripts/theme_script.R")

# load all summary data

load("./Data files/2x2/summary_results_2x2.RData")
# re-order the factor levels
summary_results <- summary_results %>%
  dplyr::mutate(scenario = as.factor(scenario),
                scenario = plyr::revalue(scenario, 
                                         c("baseline"="baseline", 
                                           "adult_missing"="adult \nmissing",
                                           "juvenile_missing" = "juvenile \nmissing",
                                           "random_error" = "random \nerror",
                                           "random_missing" = 
                                             "random \nmissing")),
                scenario = fct_relevel(scenario, 
                                       "baseline",
                                       "random \nerror",
                                       "random \nmissing",
                                       "adult \nmissing",
                                       "juvenile \nmissing"))
summary_results_2x2 <- summary_results
load("./Data files/3x3/summary_results_3x3.RData")
# re-order the factor levels
summary_results <- summary_results %>%
  dplyr::mutate(scenario = as.factor(scenario),
                scenario = plyr::revalue(scenario, 
                                         c("baseline"="baseline", 
                                           "adult_missing"="adult \nmissing",
                                           "juvenile_missing" = "juvenile \nmissing",
                                           "random_error" = "random \nerror",
                                           "random_missing" = 
                                             "random \nmissing")),
                scenario = fct_relevel(scenario, 
                                       "baseline",
                                       "random \nerror",
                                       "random \nmissing",
                                       "adult \nmissing",
                                       "juvenile \nmissing"))
summary_results_3x3 <- summary_results
load("./Data files/5x5/summary_results_5x5.RData")
# re-order the factor levels
summary_results <- summary_results %>%
  dplyr::mutate(scenario = as.factor(scenario),
                scenario = plyr::revalue(scenario, 
                                         c("baseline"="baseline", 
                                           "adult_missing"="adult \nmissing",
                                           "juvenile_missing" = "juvenile \nmissing",
                                           "random_error" = "random \nerror",
                                           "random_missing" = 
                                             "random \nmissing")),
                scenario = fct_relevel(scenario, 
                                       "baseline",
                                       "random \nerror",
                                       "random \nmissing",
                                       "adult \nmissing",
                                       "juvenile \nmissing"))
summary_results_5x5 <- summary_results

summary_results_all <- bind_rows(summary_results_2x2,
                                 summary_results_3x3,
                                 summary_results_5x5) %>%
  mutate(matrix_size = c(rep(2, 17500),
                         rep(3, 25000),
                         rep(5, 40000)))



# load all true in ci data

true_ci_2x2 <- read.csv("./Data files/2x2/true_in_ci_2x2.csv")
true_ci_3x3 <- read.csv("./Data files/3x3/true_in_ci_3x3.csv")
true_ci_5x5 <- read.csv("./Data files/5x5/true_in_ci_5x5.csv")

true_ci_all <-  bind_rows(true_ci_2x2,
                          true_ci_3x3,
                          true_ci_5x5) %>%
  mutate(matrix_size = c(rep(2, 175),
                         rep(3, 250),
                         rep(5, 400))) %>%
  dplyr::mutate(scenario = fct_relevel(scenario, 
                                       "baseline",
                                       "random \nerror",
                                       "random \nmissing",
                                       "adult \nmissing",
                                       "juvenile \nmissing"))

# DIRECT ESTIMATE FILES

full_results_2x2_direct <- read.csv("./Data files/2x2/full_direct_est_results_2x2.csv")
full_results_3x3_direct <- read.csv("./Data files/3x3/full_direct_est_results_3x3.csv")
full_results_5x5_direct <- read.csv("./Data files/5x5/full_direct_est_results_5x5.csv")

full_results_direct <- bind_rows(full_results_2x2_direct,
                                 full_results_3x3_direct,
                                 full_results_5x5_direct) %>%
  mutate(matrix_size = c(rep(2, 12500),
                         rep(3, 17500),
                         rep(5, 27500))) %>%
  dplyr::mutate(scenario = as.factor(scenario),
                scenario = fct_relevel(scenario, 
                                       "baseline",
                                       "random \nerror",
                                       "random \nmissing",
                                       "adult \nmissing",
                                       "juvenile \nmissing"))

true_ci_2x2_direct <- read.csv("./Data files/2x2/direct_est_coverage_2x2.csv")
true_ci_3x3_direct <- read.csv("./Data files/3x3/direct_est_coverage_3x3.csv")
true_ci_5x5_direct <- read.csv("./Data files/5x5/direct_est_coverage_5x5.csv")

true_ci_all_direct <- bind_rows(true_ci_2x2_direct,
                                true_ci_3x3_direct,
                                true_ci_5x5_direct) %>%
  mutate(matrix_size = c(rep(2, 125),
                         rep(3, 175),
                         rep(5, 275)))%>%
  dplyr::mutate(scenario = as.factor(scenario),
                scenario = plyr::revalue(scenario, 
                                         c("baseline"="baseline", 
                                           "adult_missing"="adult \nmissing",
                                           "juvenile_missing" = "juvenile \nmissing",
                                           "random_error" = "random \nerror",
                                           "random_missing" = 
                                             "random \nmissing")),
                scenario = fct_relevel(scenario, 
                                       "baseline",
                                       "random \nerror",
                                       "random \nmissing",
                                       "adult \nmissing",
                                       "juvenile \nmissing"))

###############################################################################

colours_scenario <- c("#FFFFFF", "#feff54", "#30666B", "#134263", "#1E2E39")

colfunc<-colorRampPalette(c("white", "orangered"))

colours_size <- colfunc(5)

```

## Introduction

* Imperfect data are common place in ecological studies. Difficulties in measuring
all individuals in a population, coupled with the cryptic nature of many wild 
organisms mean that we often cannot fully observe particular processes occurring
in natural populations. As a result, there is an observation process present in
our data and an imperfect mapping from the population of interest to our samples.
Such data must be dealt with appropriately in statistical models to prevent the 
addition of unaccounted for uncertainty into ecological results. 

* To deal with imperfect data correctly we need models that not only capture the
biological process of interest but also the observation process that gave rise to
our observed data (CITE). With recent increases in computational power and
ease of fitting more complex models on ordinary laptops (CITE), it is now accessible
to construct and fit such models more easily. Indeed, the use of these so called
state-space models in ecological studies is getting more common (CITE), but far 
from ubiquitous. 

* Matrix population models (MPMs) are widely applied ecological tools used to predict
population changes and understand various aspects of species' life histories. The
data that are the basis of these MPMs are subject to many of the difficulties
typical of ecological data; being often a small subset of the total population
and imperfectly observed, especially for elusive or cryptic species. Therefore,
the statistical models or calculations that estimate the vital rates that are the 
parameters of these models should include explicit observation processes.

* While we have a comprehensive and widely applied set of tools for account for 
the observation processes in our survival data (MARK, E-SURGE etc), the same 
sophisticated methods are rarely used for data on reproductive success. We found
in a review of published MPMs that only X% of papers modelled an explicit 
observation process for reproductive parameters, in contrast to X% for survival
parameters. 

* Therefore, observation errors in reproductive success are likely going unaccounted
for in many of our studies. Given that reproductive events can be especially hard
to identify for many species causing missed individuals and are often based on 
small sample sizes or subject to count errors (CITE) this could be a problem for 
the inference we draw from analyses using these estimates. But how much impact
would these ignored observation processes actually have on derived estimates
(such as population growth rate) from MPMs?

* In this study we explore the potential consequences of the omission of 
explicit modelling of an observation process under five scenarios of reproductive
data error and bias; perfect observation, count error, missing individuals at 
random, bias to missing first time breeders, bias to missing experienced breeders. 
All scenarios have imperfect observation of survival. 
We explore the impact of ignoring these errors and bias 
on the accuracy, precision, and precision accuracy
(coverage) of estimates of the population growth rate for three matrix sizes
and covering a range of different life histories. 

## Method

* First I selected 2x2 matrices from COMADRE (published since 2010 and in
peer reviewed literature). I reduced these to only matrices where both stage
classes breed. I then organised these based on the sum of their fecundity values.
This gave 10 matrices, so I selected every other matrix from the lowest fecundity
value to the second highest. This gave a range of different levels of fecundity.

* Next I expanded these five 2x2 matrices to 3x3 and 5x5. 

* I did this by repeating the survival and fecundity values with some tweaks.
First, the number of breeding stages varied by matrix size and was 2 for 3x3 and
4 for 5x5. Earliest breeding stage has earliest breeding stage fecundity from 
2x2 matrix, all others have repeat of adult stage fecundity. Youngest stage has
youngest stage survival from 2x2, all others repeated. 

* Stable lambda (>1 < 1.2) was needed for simulations, so if expanded matrix had
lambda < 1 all matrix elements were increased by 10%. 

* Once matrices were established, recapture data was simulated with recapture
rate of 1 for first year individuals and 0.8 thereafter. 

* Five scenarios were used

  1. **Baseline**: this has imperfect detection of survival but no error in 
  fecundity.
  
  2. **Random error**: this has imperfect detection of survival and random count
  error in fecundity - universal across all breeding groups.
  
  3. **Random missing**: this has imperfect detection of survival and imperfect
  detection of breeding events. The missing breeding events occur at random 
  (10%).
  
  4. **Missing first time breeders**: this has imperfect detection of survival
  and imperfect detection of breeding events for first time breeders (20% 
  missing).
  
  5. **Missing post-first time breeders**: this has imperfect detection of 
  survival and imperfect detection of breeding events for post-first time 
  breeders (20% missing).
  
* I then ran a Bayesian model with a hidden markov multistate model for survival
and Poisson regression for fecundity - no observation error modelled for 
fecundity. 

## Results: Accuracy

Accuracy metric: error i.e. difference between estimate and true value. 

### Accuracy - by scenario and matrix size

Accuracy organised by scenario and matrix size (2x2, 3x3, 5x5).

```{r accuracy_by_matrix_size}

# use the full data frame for plotting

ggplot(data = filter(summary_results_all, parameter == "lambda"),
       aes(x = scenario,
           y = error, fill = scenario,
           colour = scenario)) +
  geom_hline(yintercept = 0, colour = "blue", linetype = 2)+
  geom_violin(scale = "width", draw_quantiles = c(0.025, 0.5, 0.975)) +
  scale_fill_manual(values = colours_scenario) +
  scale_color_manual(values = c("black", "grey50", "white", "white", "white"))+
  plain_theme() +
  facet_grid(rows = vars(matrix_size))+
  labs(x = "", y = "Estimate - True") +
  theme(legend.position = "none")

# what if I just summarise as a mean error per scenario

mean_error_size <- summary_results_all %>%
  filter(parameter == "lambda") %>%
  group_by(scenario, matrix_size, matrix_number) %>%
  summarise(mean_error = mean(error),
            variance_error = var(error))

ggplot(data = mean_error_size,
       
       aes(y = mean_error,
           x = scenario, fill = scenario,
           colour = scenario)) +
  geom_col() +
  scale_fill_manual(values = colours_scenario) +
  scale_color_manual(values = c("black", "grey50", "grey50", "grey50", "grey50"))+
  plain_theme() +
  facet_grid(cols = vars(matrix_size),
             rows = vars(matrix_number))+
  labs(x = "", y = "Error in posterior mean estimate",
       title = "Mean accuracy error of lambda") +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45))

ggplot(data = mean_error_size,
       
       aes(y = variance_error,
           x = scenario, fill = scenario,
           colour = scenario)) +
  geom_col() +
  scale_fill_manual(values = colours_scenario) +
  scale_color_manual(values = c("black", "grey50", "grey50", "grey50", "grey50"))+
  plain_theme() +
  facet_grid(cols = vars(matrix_size),
             rows = vars(matrix_number))+
  labs(x = "", y = "Variance in error",
       title = "Variance error of lambda") +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45))

```

*Interpretation: matrix size has little impact on results, maybe slightly less
impact of bias in the larger matrix (higher role of survival).*

### Accuracy - by scenario and fecundity importance

Accuracy organised by scenario and fecundity importance (1 = lowest, 5 = highest).

Just presented for 2x2 matrices.

```{r accuracy_by_ratio}

ggplot(data = filter(summary_results_all, parameter == "lambda",
                     matrix_size == 2),
       aes(x = scenario,
           y = error, fill = scenario,
           colour = scenario)) +
  geom_hline(yintercept = 0, colour = "blue", linetype = 2)+
  geom_violin(scale = "width", draw_quantiles = c(0.025, 0.5, 0.975)) +
  scale_fill_manual(values = colours_scenario) +
  scale_color_manual(values = c("black", "grey50", "white", "white", "white"))+
  plain_theme() +
  facet_grid(rows = vars(matrix_number))+
  labs(x = "", y = "Estimate - True") +
  theme(legend.position = "none")

```

*Interpretation: Variability of error differs by matrix. Generally, there is more
error in random error scenario but this is normally distributed. Can see most
bias to estimate with random missing scenario and least for juvenile missing.*

## Results: Precision

Precision is measured by credible interval width.

### Precision - across all combos

Results shown for all matrices as all results are very similar. Little changes
CI width. 

```{r precision}

#ggplot(data = filter(summary_results_all, parameter == "lambda"),
#       aes(x = scenario,
#           y = CI_width, fill = scenario, colour = scenario)) +
#  geom_violin(scale = "width", draw_quantiles = c(0.025, 0.5, 0.975)) +
#  scale_fill_manual(values = colours_scenario) +
#  scale_color_manual(values = c("black", colours_scenario[2], "grey", "grey", "grey"))+
#  plain_theme() +
#  facet_grid(rows = vars(matrix_number), cols = vars(matrix_size))+
#  labs(x = "", y = "CI width") +
#  theme(legend.position = "none",
#        axis.text.x = element_text(angle = 45))

# think this would be better as a table of mean CI width

CI_width <- summary_results_all %>%
  group_by(matrix_size, matrix_number, scenario) %>%
  summarise(mean_CI = mean(CI_width))

ggplot(data = CI_width,
       aes(y = mean_CI,
           x = scenario, fill = scenario,
           colour = scenario)) +
  geom_col() +
  scale_fill_manual(values = colours_scenario) +
  scale_color_manual(values = c("black", "grey50", "grey50", "grey50", "grey50"))+
  plain_theme() +
  facet_grid(cols = vars(matrix_size),
             rows = vars(matrix_number))+
  labs(x = "", y = "Mean CI width",
       title = "Prescision of estimates") +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45))

```

## Results: Coverage (accuracy of precision)

Coverage is the number of times the true parameter falls in the 95% credible
interval. Would expect it to be 95% but this is not always the case. 

This gives a measure of how accurate our precision metric is. 

### Coverage - lambda by scenario

Coverage of lambda plotted with mean % for each scenario across all matrix sizes
and fecundity importance. 

```{r coverage_lambda_scenario}

# first format the data

true_ci_scenario <- true_ci_all %>%
  group_by(scenario, parameter) %>% 
  summarise(mean = mean(percent*100))

ggplot(data = filter(true_ci_scenario, parameter == "lambda"),
       aes(x = scenario,
           y = mean, fill = scenario, colour = scenario)) +
  geom_col() +
  geom_hline(yintercept = 95) +
  scale_fill_manual(values = colours_scenario) +
  scale_color_manual(values = c("black", colours_scenario[2], "grey", "grey", "grey"))+
  plain_theme() +
  #facet_grid(rows = vars(matrix_number), cols = vars(matrix_size))+
  labs(x = "", y = "Percentage true in CI") +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45)) +
  ylim(0,100)
```

*Interpretation: On average, coverage is highest in the baseline scenario but 
still not 95% expected. This is because recapture rate and consequently survival
are hard to estimate in this example due to low survival and high recapture.
Random error has little impact on coverage but all bias scenarios have a big 
impact. For this, the uncertainty estimated is way off.*

### Coverage - lambda by matrix size

Coverage of lambda plotted as mean % for matrix size across all fecundity levels.

```{r coverage_lambda_size}
# first format the data

true_ci_size <- true_ci_all %>%
  mutate(matrix_size = as.factor(matrix_size)) %>%
  group_by(matrix_size, parameter) %>% 
  summarise(mean = mean(percent*100))

ggplot(data = filter(true_ci_size, parameter == "lambda"),
       aes(x = matrix_size,
           y = mean, fill = matrix_size, colour = matrix_size)) +
  geom_col() +
  geom_hline(yintercept = 95) +
  scale_fill_manual(values = colours_size[c(1,2,5)]) +
  scale_color_manual(values = c("grey", "grey", "grey"))+
  plain_theme() +
  #facet_grid(rows = vars(matrix_number), cols = vars(matrix_size))+
  labs(x = "Matrix size", y = "Percentage true in CI") +
  theme(legend.position = "none") +
  ylim(0,100)
```

*Interpretation: As this is across all scenarios, the average % drops a lot. But
is lowest for 3x3 matrices.*

### Coverage - lambda by fecundity ratio

Coverage of lambda plotted for 2x2 matrices only and facetted by fecundity
importance (1 = lowest, 5 = highest). 

```{r coverage_lambda_ratio}

ggplot(data = filter(true_ci_all, parameter == "lambda",
                     matrix_size == 2),
       aes(x = scenario,
           y = percent*100, fill = scenario, colour = scenario)) +
  geom_col() +
  geom_hline(yintercept = 95) +
  scale_fill_manual(values = colours_scenario) +
  scale_color_manual(values = c("black", colours_scenario[2], "grey", "grey", "grey"))+
  plain_theme() +
  facet_grid(rows = vars(matrix_number))+
  labs(x = "", y = "Percentage true in CI",
       title = "lambda") +
  theme(legend.position = "none") +
  ylim(0,100)

ggplot(data = filter(true_ci_all, parameter == "survival_adult",
                     matrix_size == 2),
       aes(x = scenario,
           y = percent*100, fill = scenario, colour = scenario)) +
  geom_col() +
  geom_hline(yintercept = 95) +
  scale_fill_manual(values = colours_scenario) +
  scale_color_manual(values = c("black", colours_scenario[2], "grey", "grey", "grey"))+
  plain_theme() +
  facet_grid(rows = vars(matrix_number))+
  labs(x = "", y = "Percentage true in CI",
       title = "adult survival") +
  theme(legend.position = "none") +
  ylim(0,100)

```

*Interpretation: Coverage close to 95% for baseline but drops for all error and
bias scenarios. Suggests that uncertainty incorrectly estimated if we have
unaccounted for bias and error, especially randomly missing individuals.*

## Results: Comparison with direct estimation method

I also ran all scenarios through a model that directly estimates vital rates and
then adds uncertainty through bootstrap. This more directly follows Caswell's 
book but does not account for imperfect detection in survival.

General patterns = more error, wider confidence intervals but worse coverage.

As expected - just a sense check. 

### Direct estimate error

```{r direct_error}

#ggplot(data = filter(full_results_direct, parameter == "lambda"), 
#       aes(x = scenario, y = error, fill = scenario,
#           colour = scenario)) +
#  geom_violin(scale = "width", draw_quantiles = c(0.025, 0.5, 0.975)) +
#  scale_fill_manual(values = colours_scenario) +
#  scale_color_manual(values = c("black", "grey50", "white", "white", "white"))+
#  plain_theme() +
#  facet_grid(rows = vars(matrix_number), cols = vars(matrix_size),
#             labeller = labeller(matrix_size = c("2" = "size 2", 
#                                      "3" = "size 3",
#                                      "5" = "size 5")))+
#  labs(x = "", y = "Estimate - True") +
#  theme(legend.position = "none")

# what if I just summarise as a mean error per scenario

mean_error_size_direct <- full_results_direct %>%
  filter(parameter == "lambda") %>%
  group_by(scenario, matrix_size, matrix_number) %>%
  summarise(mean_error = mean(error),
            variance_error = var(error))

ggplot(data = mean_error_size_direct,
       aes(y = mean_error,
           x = scenario, fill = scenario,
           colour = scenario)) +
  geom_col() +
  scale_fill_manual(values = colours_scenario) +
  scale_color_manual(values = c("black", "grey50", "grey50", "grey50", "grey50"))+
  plain_theme() +
  facet_grid(cols = vars(matrix_size),
             rows = vars(matrix_number))+
  labs(x = "", y = "Error in posterior mean estimate",
       title = "Mean accuracy error of lambda") +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45)) +
  ylim(-0.1, 0.3)

ggplot(data = mean_error_size_direct,
       aes(y = variance_error,
           x = scenario, fill = scenario,
           colour = scenario)) +
  geom_col() +
  scale_fill_manual(values = colours_scenario) +
  scale_color_manual(values = c("black", "grey50", "grey50", "grey50", "grey50"))+
  plain_theme() +
  facet_grid(cols = vars(matrix_size),
             rows = vars(matrix_number))+
  labs(x = "", y = "Variance in error",
       title = "Variance error of lambda") +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45))

```

### Direct estimate CI width

```{r direct_CI_width}

ggplot(data = filter(full_results_direct, parameter == "lambda"), 
       aes(x = scenario, y = CI_width, fill = scenario,
           colour = scenario)) +
  geom_violin(scale = "width", draw_quantiles = c(0.025, 0.5, 0.975)) +
  scale_fill_manual(values = colours_scenario) +
  scale_color_manual(values = c("black", "grey50", "white", "white", "white"))+
  plain_theme() +
  facet_grid(rows = vars(matrix_number), cols = vars(matrix_size),
             labeller = labeller(matrix_size = c("2" = "size 2", 
                                      "3" = "size 3",
                                      "5" = "size 5")))+
  labs(x = "", y = "CI width") +
  theme(legend.position = "none")

```

### Direct estimate coverage

```{r direct_coverage}

ggplot(data = filter(true_ci_all_direct, parameter == "lambda"),
       aes(x = scenario,
           y = percentage, fill = scenario, colour = scenario)) +
  geom_col() +
  geom_hline(yintercept = 95) +
  scale_fill_manual(values = colours_scenario) +
  scale_color_manual(values = c("black", colours_scenario[2], "grey", "grey", "grey"))+
  plain_theme() +
  facet_grid(rows = vars(matrix_number), cols = vars(matrix_size))+
  labs(x = "", y = "Percentage true in CI",
       title = "lambda") +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45)) +
  ylim(0,100)

ggplot(data = filter(true_ci_all_direct, parameter == "survival_adult"),
       aes(x = scenario,
           y = percentage, fill = scenario, colour = scenario)) +
  geom_col() +
  geom_hline(yintercept = 95) +
  scale_fill_manual(values = colours_scenario) +
  scale_color_manual(values = c("black", colours_scenario[2], "grey", "grey", "grey"))+
  plain_theme() +
  facet_grid(rows = vars(matrix_number), cols = vars(matrix_size))+
  labs(x = "", y = "Percentage true in CI",
       title = "adult survival") +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45)) +
  ylim(0,100)

```

